---
title: "Data Visualisation with ggplot2 - Project"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# AfterWork Data Science: Data Visualisation with Python Project

## 1. Defining the Question

### a) Specifying the Data Analysis Question
Provide strategy recommendations that will lead to revenue growth.

### b) Defining the Metric for Success
The solutions to the following questions will help us answer our research question:

* When is the best time of year to book a hotel room?
* When is the optimal length of stay in order to get the best daily rate?
* How will you know if a hotel was likely to receive a disproportionately high number of special requests?

### c) Understanding the context
The dataset provided contains information on the various chain of hotels, including when customers made, the booking, length of stay, the number of adults, children, or babies, and the number of available parking spaces, among other things.
Using that data, we are supposed to offer a startegy for the hotel on revenue growth

### d) Recording the Experimental Design
By analysing data for the below questions, better strategy recommendations will be provided:

● How many bookings were canceled?
● What was the booking ratio between resort hotels and city hotels?
● What was the percentage of booking for each year?
● Which was the busiest month for hotels?
● Most guests came from which country?
● How long do most people stay in hotels?
● Which was the most booked accommodation type (Single, Couple, Family)?

### e) Data Relevance
How relevant was the provided data?

It is ok for analysis.

##Pre requisite
Let's first load the tidyverse packages for data visualisation and manipulation. Alternatively, we can load the `tidyverse` library. 

```{r}

install.packages("tidyverse")

```

```{r}
library("tidyverse")

```
## 2. Reading the Data

```{r}
# Load the data below
# --- 
# Dataset url = locally saved hotel_bookings.csv
# --- 
# YOUR CODE GOES BELOW

hotel_booking_df <- read.csv("hotel_bookings.csv")
head(hotel_booking_df)
# 
```

```{r}
# Checking the first 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 
head(hotel_booking_df, 5)
```

```{r}
# Checking the last 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 

tail(hotel_booking_df, 5)
```

```{r}
# Sample 10 rows of data
# ---
# YOUR CODE GOES BELOW
# 

sample(hotel_booking_df, 10)
```

```{r}
# Checking number of rows and columns
# ---
# YOUR CODE GOES BELOW
#  
dim(hotel_booking_df)
```

```{r}
# Checking datatypes
# ---
# YOUR CODE GOES BELOW
# 
glimpse(hotel_booking_df)
```


Record your general observations below:

There are NULL information  is agent and company
There were 118200 bookings

## 3. External Data Source Validation
The data is originally from the article Hotel Booking Demand Datasets, by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume 22, February 2019.


## 4. Data Preparation
### Performing Data Cleaning
```{r}
# Checking datatypes and missing entries of all the variables
# ---
# YOUR CODE GOES BELOW
# 
colSums(is.na(hotel_booking_df))

```

We observe the following from our dataset:

There are 4 missing data whose children is missing values

```{r}
# Checking how many duplicate rows are there in the data
# ---
# YOUR CODE GOES BELOW
# 

hotel_booking_df[duplicated(hotel_booking_df), ]
```

We choose to keep the duplicates because we don't have a unique identifier to actually proof that we have duplicates.

```{r}
# Checking if any of the columns are all null
# ---
# YOUR CODE GOES BELOW
# 

colSums(is.na(hotel_booking_df))
```

We observe the following from our dataset:

The column children has missing values, 4 of them

```{r}
# Checking if any of the rows are all null
# ---
# YOUR CODE GOES BELOW
# 

hotel_booking_df[!complete.cases(hotel_booking_df),]
```

We observe the following from our dataset:

we have 4 rows with null information under children

```{r}
# Checking the correlation of the features through the use of 
# visualizations the correlation using heatmap

# Let's first install and import the reshape2 package that will provide 
# the melt() function for creating a correlation matrix. 
# After running the following line, once prompted in the console,
# type `Yes` to install it.
install.packages("reshape2")

# loading the package
library(reshape2)
```
```{r}
head(hotel_booking_df)
```


```{r}
# Question: Create a heatmap to determine which variables are highly correlated.

# Select the desired columns
hote_booking_df2 <- hotel_booking_df %>%
  na.omit() %>%
  select(adults, children, babies, required_car_parking_spaces)
hote_booking_df2

# Create a correlation matrix
corr_matrix <- cor(hote_booking_df2, method="s")
head(corr_matrix)

# Create a table with long form
corr_df <- melt(corr_matrix)
head(corr_df)

# Plot the heatmap
ggplot(corr_df, aes(Var1, Var2, fill = value)) + 
  geom_tile(color = "black") + 
  geom_text(
    aes(label = round(value, 2)), 
    color = "white"
  ) +
  coord_fixed() + 
  labs(
    fill="booking Correlation"
  ) +
  scale_fill_gradient2(
    low = "blue", 
    high = "red",
    mid = "white", 
    limit = c(-1,1)
  ) + 
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )
```


We observe the following from our data set:

Children were related to adults most while parking were least related

```{r}
# Dropping company column because it has alot of missing values 
# and we won't need to answer any of our questions
# ---
# YOUR CODE GOES BELOW
#

hotel_booking_df = select(hotel_booking_df, -c("company"))
head(hotel_booking_df)

```

From the data variable description we see that the Distribution Channel category that tells us about Booking distribution.

The term “TA” means “Travel Agents”
The term “TO” means “Tour Operators”
This allows us to fill the missing values in the agents column with TO

```{r}
# We replace the mising values i.e. for TO
# ---
# YOUR GOES BELOW
#
```

```{r}
# We drop rows where there is no adult, baby and child as 
# these records won't help us.
# ---
# YOUR GOES BELOW
# get a new data set comprising of sum of adult, baby and child

total <- c( hotel_booking_df$adults + hotel_booking_df$children+hotel_booking_df$babies)

# filter out the records where there we no adults, baby or child

hotel_booking_clean_df<-subset(hotel_booking_df, total!= 0)
dim(hotel_booking_clean_df)
dim(hotel_booking_df)
```

```{r}
# We replace missing children values with rounded mean value
# ---
# Hint i.e. use round()
# ---
# YOUR GOES BELOW
#



hotel_booking_clean_df$children[is.na(hotel_booking_clean_df$children)] <- mean(hotel_booking_clean_df$children, na.rm = TRUE)



```

```{r}
# Checking for missing values in the dataframe
# ---
# YOUR GOES BELOW
#
colSums(is.na(hotel_booking_clean_df))
```

```{r}
# Converting the datatypes of the following columns from float to integer
# i.e. children, company, agent
# ---
# YOUR GOES BELOW
#
hotel_booking_clean_df$children <- as.integer(hotel_booking_clean_df$children)
hotel_booking_clean_df$agent <- as.integer(hotel_booking_clean_df$agent)


glimpse( hotel_booking_clean_df)
```


## 5. Solution Implementation

### 5.a) Questions

```{r}
# 1. How many bookings were cancelled?
# ---
# Visualisation: Barplot


#hotel_booking_clean_df %>% 
 # group_by(hotel) %>%
  #summarize(canceled_booking = sum(is_canceled))
cancelation_ratio_df <- hotel_booking_clean_df %>% 
   group_by(hotel) %>%
  summarise(n = n()) %>%
  mutate(is_canceled = (n/sum(n))*100)

cancelation_ratio_df

# Plotting the chart
ggplot(cancelation_ratio_df, aes(x = hotel, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "The Ratio of cancelation per hotel", 
    x = "hotel", 
    y = "cancelation_ratio"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```


```{r}
# 2. What was the booking ratio between resort hotel and city hotel?
# ---
# Barplot of booking ratio between resort hotel and city hotel

booking_ratio_df <- hotel_booking_clean_df %>% 
   group_by(hotel) %>%
  summarise(n = n()) %>%
  mutate(booking_ratio = n/sum(n))

dim(booking_ratio_df)
 


# Plotting the chart
ggplot(booking_ratio_df, aes(x = hotel, y = booking_ratio)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Booking Ratio resort hotel vs city hotel", 
    x = "hotel", 
    y = "booking_ratio"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


```{r}
# 3. What was the percentage of booking for each year?
# ---

yearly_booking_df <- hotel_booking_clean_df %>% 
   group_by(arrival_date_year) %>%
  summarise(n = n()) %>%
  mutate(percent_yearly_booking = (n/sum(n))*100)

yearly_booking_df

yearly_booking_df$arrival_date_year <- as.character(yearly_booking_df$arrival_date_year)

yearly_booking_df


# represent the data on a pie chart
ggplot(data=yearly_booking_df, aes(x="", y=percent_yearly_booking, fill=arrival_date_year)) +
  geom_bar(
    stat="identity", width=0.5, color = "white"
  ) +
  coord_polar("y") + 
  geom_text(
    aes(label = paste0(percent_yearly_booking)), position = position_stack(vjust = 0.5)
  ) +
  scale_fill_brewer(palette="Set1") +   
  labs(
    fill = "year", 
    title = "ratio"
  ) +
  theme_void()
# 
```

```{r}
# 4. Which were the most busiest months for hotels?
# ---
# 

monthly_df <- hotel_booking_clean_df %>% 
   group_by(arrival_date_month) %>%
  summarise(n = n()) %>%
  mutate(booking_ratio = n/sum(n)*100)

monthly_sort_df<-monthly_df[order(-monthly_df$booking_ratio),]

monthly_sort_df
```
## The month of August and July

```{r}
# 5. From which top 3 countries did most guests come from?
# --- 
# YOUR GOES BELOW
#
#summarize number of guests per country
countries_df <- hotel_booking_clean_df %>%
  group_by(country) %>%
  summarize(guests = sum(adults, children, babies))
#sort the countries guests in desceding order 
most_guests_per_country <- countries_df[order(-countries_df$guests),]

#select the 3 countries with the highest number of guests

head(most_guests_per_country, 3)
```

```{r}
# 6.a) How long do most people stay in hotels?
#   b) By city and resort? Separate the data by hotel

weekend_df <- hotel_booking_clean_df %>%
  group_by(hotel) %>%
  summarize(avg_guest_weekend = mean(stays_in_weekend_nights))
     #                  
week_df <- hotel_booking_clean_df %>%
  group_by(hotel) %>%
  summarize(avy_guest_week = mean(stays_in_week_nights))
duration_df <- merge(week_df,weekend_df,by="hotel")
duration_df
```



```{r}
# 7. Which was the most booked accommodation type (Single, Couple, Family)?
# ---
#

## 
```

The data was not sufficient to answer this question

### 5.b) Recommendations
From the above analysis, below are our recommendations:
1. on the month of July and August , the number of visitor is optimal and thus the hotel need to use this time to maximase profif.
2. on months like December when we expect visitor to hike, due to amny holidays, the data shows that the occupancy is usually very low, this is an area where the hotel need to capitalize by offering affordable offers.
3. city hotels have a higher booking compared to resort hotel. However, the average duration of stay is higher is resort than city hotel, in this, the hotel management need to make city hotels expensive than resort hotels since thay are for short term.



## 6. Challenging your Solution
In this step, we review our solution and implement approaches that could potentially provide a better outcome. In our case, we could propose the following question that wasn't answered in our solution because it couldn't have greatly contributed to our recommendation.

```{r}
# When should hotels provide special offers?
# offers should be when booking is at its lowest:-
monthly_df <- hotel_booking_clean_df %>% 
   group_by(arrival_date_month) %>%
  summarise(n = n()) %>%
  mutate(booking_ratio = n/sum(n)*100)

monthly_sort_df<-monthly_df[order(monthly_df$booking_ratio),]

monthly_sort_df

# 
```

Our observations:
the bookings are lowest in November, December, January and February

How does this observation tie to our solution?

The higher the bookings, the more the profit, so, by encouranging more bookings, the hotel will gain higher profit



## 7. Follow up questions
During this step, you rethink and propose other ways that you can improve your solution.

a). Did we have the right data? yes
b). Do we need other data to answer our question? yes, the room type was not clear
c). Did we have the right question? yes


